pdp: An R Package for Constructing Partial Dependence Plots

### 2017-01-03 manuscript/reviews/response.Rmd

---
title: "Response to Comments"
output:
  html_document: default
  html_notebook: default
  pdf_document: default
  word_document: default
---

Thanks to the reviewers and editor for their time and and comments. I believe that they have greatly improved, not only the manuscript, but the package it describes as well. Please find below a copy of the reviewers comments along with my responses.

@@

# -----------Reviewer 1-----------

### Comment 1.1
**Bug**: A graph is produced as expected if the code below is cut-and-pasted at the command line .  However, if the code is put into a file (say `foo.R`) and `source("foo.R")` is executed at the command line, no graph is produced.  (But note that if `plot=TRUE` is added to the call to `partial()`, then a plot is produced as expected.)
```{r, eval=FALSE}
library(randomForest)
data(airquality)
ozone.rf <- randomForest(Ozone ~ ., data=airquality, na.action=na.omit)
library(pdp)
pd <- partial(ozone.rf, pred.var="Temp")
plotPartial(pd)
```

### Response 1.1
This is simply a side effect of using `lattice` to produce the plots. If sourcing a script, then the plot needs to be explicitly printed (e.g., `print(plotPartial(pd))` or `print(partial(fit, pred.var = "x", plot = TRUE))`). This has been clarified in the package documentation and the manuscript (see **Note:** at the bottom of page 3). The fact that this worked when calling `partial` with `plot = TRUE` was a bug (which is now fixed).

### 2017-01-03 manuscript/greenwell.tex
MAJOR code cleanup!

\begin{algorithm}
\begin{enumerate}
  \item For $i \in \left\{1, 2, \dots, k\right\}$:
  \begin{enumerate}
    \item Copy the training data and replace the original values of $x_1$ with the constant $x_{1i}$.
    \item Compute the vector of predicted values from the modified copy of the training data.
    \item Compute the average prediction to obtain $\bar{f}_1\left(x_{1i}\right)$.
  \end{enumerate}
  \item Plot the pairs $\left\{x_{1i}, \bar{f}_1\left(x_{1i}\right)\right\}$ for $i = 1, 2, \dotsc, k$.
\end{enumerate}
\caption{A simple algorithm for consructing the partial dependence of the response on a single predictor $x_1$. \label{alg:pdp}}
\end{algorithm}
Algorithm~\ref{alg:pdp} can be quite computationally intensive since it involves $k$ passes over the training records. Fortunately, the algorithm can be parallelized quite easily (more on this in Section~\ref{sec:computational}). It can also be easily extended to larger subsets of two or more features as well.

@@

The \CRANpkg{plotmo} package \citep{plotmo-pkg} is one alternative to \pkg{pdp}. According to \citeauthor{plotmo-pkg}, \pkg{plotmo} constructs "a poor man's partial dependence plot." In particular, it plots a model's response when varying one or two predictors while holding the other predictors in the model constant (continuous features are fixed at their median value, while factors are held at their first level). These plots allow for up to two variables at a time. They are also less accurate than PDPs, but are faster to construct. For additive models (i.e., models with no interactions), these plots are identical in shape to PDPs. As of \pkg{plotmo} version 3.3.0, there is now support for constructing PDPs, but it is not the default. The main difference is that \pkg{plotmo}, rather than applying step 1. (a)-(c) in Algorithm~\ref{alg:pdp}, accumulates all the data at once thereby reducing the number of internal calls to \code{predict}. The trade-off is a slight increase in speed at the expense of using more memory. So, why use the \pkg{pdp} package? As will be discussed in the upcoming sections, \pkg{pdp}:
\begin{itemize}
  \item contains only a few functions with relatively few arguments;
  \item does \underline{not} produce a plot by default;
  \item can be used more efficiently with \code{"gbm"} objects (see Section~\ref{sec:computational});
  \item produces graphics based on \CRANpkg{lattice} \citep{lattice-pkg}, which are far more flexible than base R graphcs;
  \item defaults to using false color level plots for multivariate displays (see Section~\ref{sec:multi});
  \item contains options to mitigate the risks associated with extrapolation (see Section~\ref{sec:computational});
  \item has the option to construct PDPs in parallel (see Section~\ref{sec:computational});
    \item is extremely flexible in the types of PDPs that can be produced (see Section~\ref{sec:prediction}),
\end{itemize}

@@

Many other techniques exist for visualizing relationships between the predictors and the response based on a fitted model. For example, the \CRANpkg{car} package \citep{fox-car-2011} contains many functions for constructing \dfn{partial-residual} and \dfn{marginal-model} plots. \dfn{Effect displays}, available in the \CRANpkg{effects} package \citep{fox-effects-2003}, provides tabular and graphical displays for the terms in parametric models while holding all other predcitors at some constant value---similar in spirit to \pkg{plotmo}'s marginal model plots. However, these methods were designed for simpler parametric models (e.g., linear and generalized linear models), whereas \pkg{plotmo}, \pkg{ICEbox}, and \pkg{pdp} are more useful for black box models (although, they can be used for simple parametric models as well).

@@

The \code{partial} function evaluates the partial dependence \eqref{eqn:pdf} from a fitted model over a grid of predictor values; the fitted model and predictors are specified using the \code{object} and \code{pred.var} arguments, respectively---these are the only required arguments. If \code{plot = FALSE} (the default), \code{partial} returns an object of class \code{"partial"} which inherits from the class \code{"data.frame"}; put another way, by default, \code{partial} returns a data frame with an additional class that is recognized by the \code{plotPartial} function. The columns of the data frame are labeled in the same order as the features supplied to \code{pred.var}, and the last column is labeled \code{yhat}\footnote{There is one exception to this. When a function supplied via the \code{pred.fun} argument returns multiple predictions, the second to last and last columns will be labeled \code{yhat} and \code{yhat.id}, respectively (see Section~\ref{sec:prediction}).} and contains the values of the partial dependence function $\bar{f}_s\left(\boldsymbol{z}_s\right)$. If \code{plot = TRUE}, then \code{partial} makes an internal call to \code{plotPartial} (with fewer plotting options) and returns the PDP in the form of a \pkg{lattice} plot (i.e., a \code{"trellis"} object). \strong{Note:}  it is reccomended to call \code{partial} with \code{plot = FALSE} and store the results; this allows for more flexible plotting, and the user will not have to waste time calling \code{partial} again if the default plot is not sufficient.

@@

Another important argument is \code{train}. If \code{train = NULL} (the default), \code{partial} tries to extract the original training data from the fitted model object. For objects that typically store a copy of the training data (e.g., objects of class \code{"BinaryTree"}, \code{"RandomForest"}, and \code{"train"}), this is straightforward. Otherwise, \code{partial} will attempt to extract the call stored in \code{object} (if available) and use that to evaluate the training data in the same environment from which \code{partial} was called. This can cause problems when, for example, the training data have been changed after fitting the model, but before calling \code{partial}. Hence, it is good practice to always supply the training data via the \code{train} argument in the call to \code{partial}\footnote{For brevity, we ignore this option in most of the examples in this paper.}. If \code{train = NULL} and the training data can not be extracted from the fitted model, the user will be prompted with an informative error message (this will occur, for example, when using \code{partial} with \code{"ksvm"} and \code{"xgb.Booster"} objects):
\begin{example}
Error: The training data could not be extracted from object. Please supply
the raw training data using the `train` argument in the call to `partial`.
\end{example}

@@

For illustration, we will use the (corrected) Boston housing data which are available from the \CRANpkg{mlbench} package \citep{mlbench-pkg}. These data contain the median value of owner-occupied homes in 506 U.S. census tracts in the Boston area, along with 13 independent variables such as the per capita crime rate by town. These are the same data analyzed in \citet{harrison-1978-hedonic}, but with a few corrections and additional spatial variables. We begin by loading the data and omitting unimportant columns.
\begin{example}
data(BostonHousing2, package = "mlbench")  # mlbench must be installed!
boston <- BostonHousing2[, -c(1, 2, 5)]
\end{example}
Next, we fit a random forest to the entire data set with default tuning parameters and 500 trees:
\begin{example}
library(randomForest)
set.seed(101)  # for reproducibility
boston.rf <- randomForest(cmedv ~ ., data = boston, importance = TRUE)
varImpPlot(boston.rf)  # Figure 1
\end{example}
The model fit is reasonable, with an \dfn{out-of-bag} (pseudo) $R^2$ of 0.89. The variable importance scores are displayed in Figure~\ref{fig:plotmo_vs_partial}. Both plots indicate that the percentage of lower status of the population (\code{lstat}) and the average number of rooms per dwelling (\code{rm}) are highly associated with the median value of owner-occupied homes (\code{cmedv}). The question then arises, "What is the nature of these associations?" To help answer this, we can look at the partial dependence of \code{cmedv} on \code{lstat} and \code{rm}, both individually and together.

@@

As previously mentioned, the \code{randomForest} package has its own \code{partialPlot} function for visualizing the partial dependence of the response on a single predictor---the keywords here are "single predictor". For example, the following snippet of code plots the partial dependence of \code{cmedv} on \code{lstat}:
\begin{example}
partialPlot(boston.rf, pred.data = boston, x.var = "lstat")
\end{example}
The same plot can be achieved using the \code{partial} function and setting \code{plot = TRUE} (see the left side of Figure~\ref{fig:pd_lstat}):
\begin{example}
library(pdp)
partial(boston.rf, pred.var = "lstat", plot = TRUE)  # Figure 2 (left)
\end{example}
The only difference is that \pkg{pdp} uses the \pkg{lattice} graphics package to produce all of its displays.

@@

Constructing PDPs can be quite computationally expensive\footnote{The exception is regression trees based on single-variable splits which can make use of the efficient weighted tree traversal method described in \citet{friedman-2001-greedy}, however, only the \pkg{gbm} package seems to make use of this approach; consequently, \pkg{pdp} can also exploit this strategy when used with \pkg{gbm} models (see \code{?partial} for details).} Several strategies are available to ease the computational burden in larger problems. For example, there is no need to compute partial dependence of \code{cmedv} using each unique value of \code{rm} in the training data (which would require $k = 446$ passes over the data!). We could get very reasonable results using a reduced number of points. Current options are to use a grid of equally spaced values in the range of the variable of interest; the number of points can be controlled using the \code{grid.resolution} option in the call to \code{partial}. Alternatively, a user-specified grid of values (e.g., containing specific quantiles of interest) can be supplied through the \code{pred.grid} argument. To demonstrate, the following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} using each option, then uses \code{grid.arrange} to display all three PDPs on the same page; the results are displayed in Figure~\ref{fig:partial_manual}.
\begin{example}
# Display plots side by side (Figure 5)
grid.arrange(
  partial(boston.rf, "rm", plot = TRUE),
  partial(boston.rf, "rm", grid.resolution = 30, plot = TRUE),
  partial(boston.rf, "rm", pred.grid = data.frame(rm = 3:9), plot = TRUE),
  ncol = 3
)
\end{example}

@@

Next, we use the multivariate adaptive regression splines (MARS) algorithm introduced in \citet{friedman-1991-mars} to model ozone concentration as a nonlinear function of the eight meteorological variables plus day of the year; we allow for up to three-way interactions.
\begin{example}
ozone.mars <- earth(ozone ~ ., data = ozone, degree = 3)
summary(ozone.mars)
\end{example}
The MARS model produced a generalized $R^2$ of $0.79$, similar to what was reported in \citet{breiman-1985-estimating}. A single three-way interaction was found involving the predictors
\begin{itemize}
  \item \code{wind}: wind speed (mph) at Los Angeles International Airport (LAX)
  \item \code{temp}: temperature ($^oF$) at Sandburg Air Force Base
  \item \code{dpg}: the pressure gradient (mm Hg) from LAX to Dagget, CA
\end{itemize}
To understand this interaction, we can use a PDP. However, since the partial dependence between three continuous variables can be computationally expensive, we will run \code{partial} in parallel.

@@

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=0.8\linewidth]{partial_par}
  \caption{Partial dependence of \code{ozone} on \code{wind}, \code{temp}, and \code{dpg}. Since \code{dpg} is contiuous, it is first converted to a shingle; in this case, four groups with 10\% overlap.}
  \label{fig:partial_par}
\end{figure}

@@

To illustrate, we consider Edgar Anderson's iris data from the \pkg{datasets} package. The \code{iris} data frame contains the sepal length, sepal width, petal length, and petal width (in centimeters) for 50 flowers from each of three species of iris:  setosa, versicolor, and virginica. We fit a support vector machine with a Gaussian radial basis function kernel to the data using the \code{svm} function in the \pkg{e1071} package (the tuning parameters were determined using 5-fold cross-validation).
\begin{example}
library(e1071)
iris.svm <- svm(Species ~ ., data = iris, kernel = "radial", gamma = 0.75,
                cost = 0.25, probability = TRUE)
\end{example}
\strong{Note:} the \code{partial} function has to be able to extract the predicted probabilities for each class, so it is necessary to set \code{probability = TRUE} in the call to \code{svm}.

Next, we plot the partial dependence of \code{Species} on both \code{Petal.Width} and \code{Petal.Length} for each of the three classes. The result is displayed in Figure~\ref{fig:partial_iris}. %Note that we had to supply the original training data to \code{partial} through the \code{train} option. This is necessary for fitted model objects that don't store a copy of the training data (e.g., model fitting functions that don't rely on a formula method).
\begin{example}
pd <- NULL
for (i in 1:3) {
  tmp <- partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"),
                 which.class = i, grid.resolution = 101, progress = "text")
  pd <- rbind(pd, cbind(tmp, Species = levels(iris$Species)[i]))
}

@@

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{partial_iris_svm}
  \caption{Partial dependence of species on petal width and petal length for the iris data.}
  \label{fig:partial_iris}
\end{figure}

@@

With \code{partial} it is possible to replace the mean in step 1. (c) of Algorithm~\ref{alg:pdp} with any other function (e.g., the median or trimmed mean), or obtain PDPs for classification problems on the probability scale. It is even possible to obtain ICE curves. This flexibility is due to the new \code{pred.fun} argument in \code{partial} (starting with \pkg{pdp} version 0.4.0). This argument accepts an optional prediction function that requires two arguments: \code{object} and \code{newdata}. The supplied prediction function must return either a single prediction or a vector of predictions. Returing the mean of all the predictions will result in the traditional PDP. Returing a vector of predictions (i.e., one for each observation) will result in a set of ICE curves. The examples below illustrate.

@@

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{partial_iris_svm_prob}
  \caption{Partial dependence of \code{Species} on \code{Petal.Width} and \code{Petal.Length} plotted on the probability scale; in this case, the probabiltity of belonging to the \code{Setosa} species.}
  \label{fig:partial_iris_prob}
\end{figure}

@@

The curves in Figure~\ref{fig:ice_boston} indicate some heterogeneity in the fitted model (i.e., some of the curves depict the opposite relationship). Such heterogeneity can be easier to spot using c-ICE curves; see Equation (4) on page 49 of \citet{goldstein-peeking-2015}. Using \CRANpkg{dplyr} \citep{dplyr-pkg}, it is rather straightforward to post-process the output from \code{partial} to obtain c-ICE curves (similar to the construction of \dfn{raw change scores} \citep[pg. 130]{fitzmaurice-2011-applied} for longitudinal data). This is shown below.
\begin{example}
# Post-process rm.ice to obtain c-ICE curves
library(dplyr)
rm.ice <- rm.ice %>%
  group_by(yhat.id) %>%
  mutate(yhat.centered = yhat - first(yhat))
\end{example}

@@

The \code{train} function creates objects of class \code{"train"}, whereas the \code{xgboost} function creates objects of class \code{"xgb.Booster"}. Since \code{train} defaults to storing a copy of the training data as part of the \code{"train"} object, there is no need to supply it in the call to \code{partial} in this example. However, this is not the case when using the \pkg{xgboost} package directly. To illustrate, we fit the same model using the \code{xgboost} function with the tuning parameters found previously using \pkg{caret}.
\begin{example}
set.seed(203)  # for reproducibility
boston.xgb <- xgboost(data = data.matrix(subset(boston, select = -cmedv)),
                      label = boston$cmedv, objective = "reg:linear",
                      nrounds = 100, max_depth = 5, eta = 0.3, gamma = 0,
                      colsample_bytree = 0.8, min_child_weight = 1,
                      subsample = 0.9444444)
\end{example}
To use \code{partial} with \code{"xgb.Booster"} objects, we need to supply the original training data (minus the response) in the call to \code{partial}. The following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} (plot not shown). (Make sure you are using version 0.6-0 or later of \pkg{xgboost}: \url{https://github.com/dmlc/xgboost/tree/master/R-package}.) \strong{Note:} while \code{xgboost} requires the training data to be an object of class \code{"matrix"}, \code{"dgCMatrix"}, or \code{"xgb.DMatrix"}, \code{partial} requires a \code{"data.frame"} that does not contain the response column.
\begin{example}
partial(boston.xgb, pred.var = "rm", plot = TRUE, rug = TRUE,
        train = subset(boston, select = -cmedv))
\end{example}

>>>>

\begin{algorithm}
\begin{enumerate}
  \item For $i \in \left\{1, 2, \dots, k\right\}$:
  \begin{enumerate}
    \item Copy the training data and replace the original values of $x_1$ with the constant $x_{1i}$.
    \item Compute the vector of predicted values from the modified copy of the training data.
    \item Compute the average prediction to obtain $\bar{f}_1\left(x_{1i}\right)$.
  \end{enumerate}
  \item Plot the pairs $\left\{x_{1i}, \bar{f}_1\left(x_{1i}\right)\right\}$ for $i = 1, 2, \dotsc, k$.
\end{enumerate}
\caption{A simple algorithm for constructing the partial dependence of the response on a single predictor $x_1$. \label{alg:pdp}}
\end{algorithm}
Algorithm~\ref{alg:pdp} can be quite computationally intensive since it involves $k$ passes over the training records. Fortunately, the algorithm can be parallelized quite easily (more on this in Section~\ref{sec:computational}). It can also be easily extended to larger subsets of two or more features as well.

@@

The \CRANpkg{plotmo} package \citep{plotmo-pkg} is one alternative to \pkg{pdp}. According to \citeauthor{plotmo-pkg}, \pkg{plotmo} constructs "a poor man's partial dependence plot." In particular, it plots a model's response when varying one or two predictors while holding the other predictors in the model constant (continuous features are fixed at their median value, while factors are held at their first level). These plots allow for up to two variables at a time. They are also less accurate than PDPs, but are faster to construct. For additive models (i.e., models with no interactions), these plots are identical in shape to PDPs. As of \pkg{plotmo} version 3.3.0, there is now support for constructing PDPs, but it is not the default. The main difference is that \pkg{plotmo}, rather than applying step 1. (a)-(c) in Algorithm~\ref{alg:pdp}, accumulates all the data at once thereby reducing the number of internal calls to \code{predict}. The trade-off is a slight increase in speed at the expense of using more memory. So, why use the \pkg{pdp} package? As will be discussed in the upcoming sections, \pkg{pdp}:
\begin{itemize}
  \item contains only a few functions with relatively few arguments;
  \item does \underline{not} produce a plot by default;
  \item can be used more efficiently with \code{"gbm"} objects (see Section~\ref{sec:computational});
  \item produces graphics based on \CRANpkg{lattice} \citep{lattice-pkg}, which are far more flexible than base R graphics;
  \item defaults to using false color level plots for multivariate displays (see Section~\ref{sec:multi});
  \item contains options to mitigate the risks associated with extrapolation (see Section~\ref{sec:computational});
  \item has the option to construct PDPs in parallel (see Section~\ref{sec:computational});
    \item is extremely flexible in the types of PDPs that can be produced (see Section~\ref{sec:prediction}),
\end{itemize}

@@

Many other techniques exist for visualizing relationships between the predictors and the response based on a fitted model. For example, the \CRANpkg{car} package \citep{fox-car-2011} contains many functions for constructing \dfn{partial-residual} and \dfn{marginal-model} plots. \dfn{Effect displays}, available in the \CRANpkg{effects} package \citep{fox-effects-2003}, provides tabular and graphical displays for the terms in parametric models while holding all other predictors at some constant value---similar in spirit to \pkg{plotmo}'s marginal model plots. However, these methods were designed for simpler parametric models (e.g., linear and generalized linear models), whereas \pkg{plotmo}, \pkg{ICEbox}, and \pkg{pdp} are more useful for black box models (although, they can be used for simple parametric models as well).

@@

The \code{partial} function evaluates the partial dependence \eqref{eqn:pdf} from a fitted model over a grid of predictor values; the fitted model and predictors are specified using the \code{object} and \code{pred.var} arguments, respectively---these are the only required arguments. If \code{plot = FALSE} (the default), \code{partial} returns an object of class \code{"partial"} which inherits from the class \code{"data.frame"}; put another way, by default, \code{partial} returns a data frame with an additional class that is recognized by the \code{plotPartial} function. The columns of the data frame are labeled in the same order as the features supplied to \code{pred.var}, and the last column is labeled \code{yhat}\footnote{There is one exception to this. When a function supplied via the \code{pred.fun} argument returns multiple predictions, the second to last and last columns will be labeled \code{yhat} and \code{yhat.id}, respectively (see Section~\ref{sec:prediction}).} and contains the values of the partial dependence function $\bar{f}_s\left(\boldsymbol{z}_s\right)$. If \code{plot = TRUE}, then \code{partial} makes an internal call to \code{plotPartial} (with fewer plotting options) and returns the PDP in the form of a \pkg{lattice} plot (i.e., a \code{"trellis"} object). \strong{Note:}  it is recommended to call \code{partial} with \code{plot = FALSE} and store the results; this allows for more flexible plotting, and the user will not have to waste time calling \code{partial} again if the default plot is not sufficient.

@@

@@

Another important argument to \code{partial} is \code{train}. If \code{train = NULL} (the default), \code{partial} tries to extract the original training data from the fitted model object. For objects that typically store a copy of the training data (e.g., objects of class \code{"BinaryTree"}, \code{"RandomForest"}, and \code{"train"}), this is straightforward. Otherwise, \code{partial} will attempt to extract the call stored in \code{object} (if available) and use that to evaluate the training data in the same environment from which \code{partial} was called. This can cause problems when, for example, the training data have been changed after fitting the model, but before calling \code{partial}. Hence, it is good practice to always supply the training data via the \code{train} argument in the call to \code{partial}\footnote{For brevity, we ignore this option in most of the examples in this paper.}. If \code{train = NULL} and the training data can not be extracted from the fitted model, the user will be prompted with an informative error message (this will occur, for example, when using \code{partial} with \code{"ksvm"} and \code{"xgb.Booster"} objects):
\begin{example}
Error: The training data could not be extracted from object. Please supply
the raw training data using the `train` argument in the call to `partial`.
\end{example}

For illustration, we will use the (corrected) Boston housing data which are available from the \CRANpkg{mlbench} package \citep{mlbench-pkg}. These data contain the median value of owner-occupied homes in 506 U.S. census tracts in the Boston area, along with 13 independent variables such as the per capita crime rate by town. These are the same data analyzed in \citet{harrison-1978-hedonic}, but with a few corrections and additional spatial variables. We begin by loading the data and omitting unimportant columns.
\begin{example}
data(BostonHousing2, package = "mlbench")  # mlbench must be installed!
boston <- BostonHousing2[, -c(1, 2, 5)]
\end{example}
Next, we fit a random forest to the entire data set with default tuning parameters and 500 trees:
\begin{example}
library(randomForest)  # for randomForest, partialPlot, and varImpPlot functions
set.seed(101)  # for reproducibility
boston.rf <- randomForest(cmedv ~ ., data = boston, importance = TRUE)
varImpPlot(boston.rf)  # Figure 1
\end{example}
The model fit is reasonable, with an \dfn{out-of-bag} (pseudo) $R^2$ of 0.89. The variable importance scores are displayed in Figure~\ref{fig:plotmo_vs_partial}. Both plots indicate that the percentage of lower status of the population (\code{lstat}) and the average number of rooms per dwelling (\code{rm}) are highly associated with the median value of owner-occupied homes (\code{cmedv}). The question then arises, "What is the nature of these associations?" To help answer this, we can look at the partial dependence of \code{cmedv} on \code{lstat} and \code{rm}, both individually and together.

@@

As previously mentioned, the \code{randomForest} package has its own \code{partialPlot} function for visualizing the partial dependence of the response on a single predictor---the keywords here are "single predictor". For example, the following snippet of code plots the partial dependence of \code{cmedv} on \code{lstat}:
\begin{example}
partialPlot(boston.rf, pred.data = boston, x.var = "lstat")
\end{example}
The same plot can be achieved using the \code{partial} function and setting \code{plot = TRUE} (see the left side of Figure~\ref{fig:pd_lstat}):
\begin{example}
library(pdp)  # for partial, plotPartial, and grid.arrange functions
partial(boston.rf, pred.var = "lstat", plot = TRUE)  # Figure 2 (left)
\end{example}
The only difference is that \pkg{pdp} uses the \pkg{lattice} graphics package to produce all of its displays.

@@

Constructing PDPs can be quite computationally expensive\footnote{The exception is regression trees based on single-variable splits which can make use of the efficient weighted tree traversal method described in \citet{friedman-2001-greedy}, however, only the \pkg{gbm} package seems to make use of this approach; consequently, \pkg{pdp} can also exploit this strategy when used with \pkg{gbm} models (see \code{?partial} for details).} Several strategies are available to ease the computational burden in larger problems. For example, there is no need to compute partial dependence of \code{cmedv} using each unique value of \code{rm} in the training data (which would require $k = 446$ passes over the data!). We could get very reasonable results using a reduced number of points. Current options are to use a grid of equally spaced values in the range of the variable of interest; the number of points can be controlled using the \code{grid.resolution} option in the call to \code{partial}. Alternatively, a user-specified grid of values (e.g., containing specific quantiles of interest) can be supplied through the \code{pred.grid} argument. To demonstrate, the following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} using each option; \code{grid.arrange} is used to display all three PDPs on the same graph, side by side. The results are displayed in Figure~\ref{fig:partial_manual}.
\begin{example}
# Figure 5
grid.arrange(
  partial(boston.rf, "rm", plot = TRUE),
  partial(boston.rf, "rm", grid.resolution = 30, plot = TRUE),
  partial(boston.rf, "rm", pred.grid = data.frame(rm = 3:9), plot = TRUE),
  ncol = 3
)
\end{example}

@@

Next, we use the multivariate adaptive regression splines (MARS) algorithm introduced in \citet{friedman-1991-mars} to model ozone concentration as a nonlinear function of the eight meteorological variables plus day of the year; we allow for up to three-way interactions.
\begin{example}
library(earth)  # for earth function (i.e., MARS algorithm)
ozone.mars <- earth(ozone ~ ., data = ozone, degree = 3)
summary(ozone.mars)
\end{example}
The MARS model produced a generalized $R^2$ of $0.79$, similar to what was reported in \citet{breiman-1985-estimating}. A single three-way interaction was found involving the predictors
\begin{itemize}
  \item \code{wind}: wind speed (mph) at Los Angeles International Airport (LAX)
  \item \code{temp}: temperature ($^oF$) at Sandburg Air Force Base
  \item \code{dpg}: the pressure gradient (mm Hg) from LAX to Dagget, CA
\end{itemize}
To understand this interaction, we can use a PDP. However, since the partial dependence between three continuous variables can be computationally expensive, we will run \code{partial} in parallel.

@@

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=0.8\linewidth]{partial_par}
  \caption{Partial dependence of \code{ozone} on \code{wind}, \code{temp}, and \code{dpg}. Since \code{dpg} is continuous, it is first converted to a shingle; in this case, four groups with 10\% overlap.}
  \label{fig:partial_par}
\end{figure}

@@

To illustrate, we consider Edgar Anderson's iris data from the \pkg{datasets} package. The \code{iris} data frame contains the sepal length, sepal width, petal length, and petal width (in centimeters) for 50 flowers from each of three species of iris: setosa, versicolor, and virginica. We fit a support vector machine with a Gaussian radial basis function kernel to the data using the \code{svm} function in the \pkg{e1071} package (the tuning parameters were determined using 5-fold cross-validation).
\begin{example}
library(e1071)  # for svm function
iris.svm <- svm(Species ~ ., data = iris, kernel = "radial", gamma = 0.75,
                cost = 0.25, probability = TRUE)
\end{example}
\strong{Note:} the \code{partial} function has to be able to extract the predicted probabilities for each class, so it is necessary to set \code{probability = TRUE} in the call to \code{svm}.

Next, we plot the partial dependence of \code{Species} on both \code{Petal.Width} and \code{Petal.Length} for each of the three classes. The result is displayed in Figure~\ref{fig:partial_iris}.
\begin{example}
pd <- NULL
for (i in 1:3) {
  tmp <- partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"),
                 which.class = i, grid.resolution = 101, progress = "text")
  pd <- rbind(pd, cbind(tmp, Species = levels(iris$Species)[i]))
}

@@

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{partial_iris_svm}
  \caption{Partial dependence of \code{Species} on \code{Petal.Width} and \code{Petal.Length} for the iris data.}
  \label{fig:partial_iris}
\end{figure}

@@

With \code{partial} it is possible to replace the mean in step 1. (c) of Algorithm~\ref{alg:pdp} with any other function (e.g., the median or trimmed mean), or obtain PDPs for classification problems on the probability scale. It is even possible to obtain ICE curves. This flexibility is due to the new \code{pred.fun} argument in \code{partial} (starting with \pkg{pdp} version 0.4.0). This argument accepts an optional prediction function that requires two arguments: \code{object} and \code{newdata}. The supplied prediction function must return either a single prediction or a vector of predictions. Returning the mean of all the predictions will result in the traditional PDP. Returning a vector of predictions (i.e., one for each observation) will result in a set of ICE curves. The examples below illustrate.

@@

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{partial_iris_svm_prob}
  \caption{Partial dependence of \code{Species} on \code{Petal.Width} and \code{Petal.Length} plotted on the probability scale; in this case, the probability of belonging to the setosa species.}
  \label{fig:partial_iris_prob}
\end{figure}

@@

The curves in Figure~\ref{fig:ice_boston} indicate some heterogeneity in the fitted model (i.e., some of the curves depict the opposite relationship). Such heterogeneity can be easier to spot using c-ICE curves; see Equation (4) on page 49 of \citet{goldstein-peeking-2015}. Using \CRANpkg{dplyr} \citep{dplyr-pkg}, it is rather straightforward to post-process the output from \code{partial} to obtain c-ICE curves (similar to the construction of \dfn{raw change scores} \citep[pg. 130]{fitzmaurice-2011-applied} for longitudinal data). This is shown below.
\begin{example}
# Post-process rm.ice to obtain c-ICE curves
library(dplyr)  # for group_by and mutate functions
rm.ice <- rm.ice %>%
  group_by(yhat.id) %>%
  mutate(yhat.centered = yhat - first(yhat))
\end{example}

@@

The \code{train} function creates objects of class \code{"train"}, whereas the \code{xgboost} function creates objects of class \code{"xgb.Booster"}. Since \code{train} defaults to storing a copy of the training data as part of the \code{"train"} object, there is no need to supply it in the call to \code{partial} in this example. However, this is not the case when using the \pkg{xgboost} package directly. To illustrate, we fit the same model using the \code{xgboost} function with the tuning parameters found previously using \pkg{caret}.
\begin{example}
library(xgboost)  # for xgboost function
set.seed(203)  # for reproducibility
boston.xgb <- xgboost(data = data.matrix(subset(boston, select = -cmedv)),
                      label = boston$cmedv, objective = "reg:linear",
                      nrounds = 100, max_depth = 5, eta = 0.3, gamma = 0,
                      colsample_bytree = 0.8, min_child_weight = 1,
                      subsample = 0.9444444)
\end{example}
To use \code{partial} with \code{"xgb.Booster"} objects, we need to supply the original training data (minus the response) in the call to \code{partial}. The following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} (plot not shown). (Make sure you are using version 0.6-0 or later of \pkg{xgboost}: \url{https://github.com/dmlc/xgboost/tree/master/R-package}.) \strong{Note:} while \code{xgboost} requires the training data to be an object of class \code{"matrix"}, \code{"dgCMatrix"}, or \code{"xgb.DMatrix"}, \code{partial} requires a \code{"data.frame"} that does not contain the response column.
\begin{example}
partial(boston.xgb, pred.var = "rm", plot = TRUE, rug = TRUE,
        train = subset(boston, select = -cmedv))
\end{example}
